---
title: "Dycone erythrocite example"
author: "Christian Diener"
date: "`r Sys.Date()`"
output:
  html_document:
    highlight: default
    theme: default
vignette: >
  %\VignetteIndexEntry{Dycone erythrocite example}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

Human blood cell example
========================

In this example we will use dycone to analyze some perturbations in the 
Red blood cell model of Resendis. First we will use dycone to get its 
stochiometric matrix and extract the steady state from the orginial SBML file.

```{r}
library(dycone)
data(eryth)
reacts = make_irreversible(eryth)
print(eryth)

concs = read.csv( system.file("extdata", "eryth_ic.csv", package="dycone") )
ids = concs$id
concs = concs$initial
names(concs) = ids
print(concs)
```

We will treat species with lacking steady state concentrations as constant.
```{r}
spec = species(eryth)
no_c = !(spec %in% ids)
constants = spec[no_c]
concs = concs[spec[!no_c]]
```

We extract the stochiometric matrix with the constant species simply by:
```{r}
S = stochiometry(eryth, const=constants)
```

Extracting the k-cone
---------------------

Getting the consistent k-cone (all $k_i>0$) takes two steps in dycone. First, 
calculation of the general flux cone and, second, calculation of the specific
flux cone. Let's calculate and look at the flux cone first...

```{r fig.width=16, fig.height=8}
V = polytope_basis(S)
plot_basis(V)
```

Some interesting structurs can be observed from the flux kone alone. For instance, 
basis vectors with only two reactions indicate a constant relation of the form
$k_i/k_j = const.$.

Perturbance experiment
----------------------

Now, let's set up a small perturbation experiment in order to execute the 
differential k-cone analysis. We will use a faster version of the eigendynamics
as reference parameters and perturb the steady state value of pyruvate.

```{r}
m = ma_terms(S, concs)
K = kcone(V, m)
k = 1e3*eigendynamics(K)

# Now we generate several new steady states coming from a disturbed ethanol 
# production 

scales = 10^c(-1,-2)
d_concs = concs
for( i in 1:length(scales) ) {
	d_c = concs
	d_c["pyr"] = d_c["pyr"]*scales[i]
	d_concs = rbind(d_concs, d_c)
}

m_terms = apply(d_concs, 1, function(co) ma_terms(S, co))
d_basis = lapply(1:3, function(i) kcone(V, m_terms[,i]))
``` 

First we will have a look at the projection of the respective k-cones in a 2D space
with colors from green to red indicating a diminishing ethanol production.

```{r fig.width=6, fig.height=6}
plot_red(d_basis)
```

This gives a visual representation but no detailed information which reactions 
are the most likely to be perturbed in the transition from the high to low
pyruvate state. This can be achieved by the `hyp` function. We will add noise
samples to each condition to simulate an experimental measurement.

```{r}
noise = function(x) {
    new = pmax(0, x*rnorm(length(x),1,0.25))
    names(new) = names(x)
    return(new)
}

set.seed(42)

exp_concs = cbind(noise(d_concs[1,]), noise)
normal = replicate(4, ma_terms(S, noise(d_concs[1,])))
disease = replicate(4, ma_terms(S, noise(d_concs[3,])))

h = hyp(normal, disease, reacts)
print(h[h$pval<0.05,])
```

We see that the most prominent change is an increased reaction rate for two reactions, 
which result to be the two reactions consuming Pyruvate. However, there are also some other
changes, however, their fold_changes are much smaller than 2, indicating only slight alterations
due to the added noise. The mean log-fold change of the alterations in pyruvate
production are close to the actual alteration (96-fold predicted vs 100-fold truth).

As we can see the change in Pyruvate is explained by either up-regulating the pyruvate
consumption or (less likely) a lower production of upstream substrates which might have to act together. 
As we can see the reaction producing pyruvate is not in the list. This makes sense since it is a 
unidirectional reaction whose down-regulation would affect the steady state quite severely
in the concentration of adp and pep, but not pyruvate, thus making it an unlikely target 
when observing a lower pyruvate concentration alone.
