---
title: "Dycone erythrocite example"
author: "Christian Diener"
date: "`r Sys.Date()`"
output:
  html_document:
    highlight: default
    theme: flatly
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

Human blood cell example
========================

In this example we will use dycone to analyze some perturbations in the 
Red blood cell model of Resendis. First we will use dycone to get its stochiometric matrix and extract the steady state from the orginial SBML file.

```{r}
library(dycone)
data(eryth)
reacts = make_irreversible(eryth)
print(eryth)

concs = read.csv("../stuff/eryth_ic.csv")
ids = concs$id
concs = concs$initial
names(concs) = ids
print(concs)
```

We will treat species with lacking steady state concentrations as constant.
```{r}
species = get_species(eryth)
no_c = !(species %in% ids)
constants = species[no_c]
concs = concs[species[!no_c]]
```

We extract the stochiometric matrix with the constant species simply by:
```{r}
S = get_stochiometry(eryth, const=constants)
```

Extracting the k-cone
---------------------

Getting the consistent (all $k_i>0$) is simple with dycone and can be done straight
forward. Let's also extract the eigenpathway at once.

```{r fig.width=24, fig.height=8}
kc = get_polytope_basis(S, get_ma_terms(S, concs))
ep = eigenpathways(kc)[,1]
plot(kc, key=F)
```

Some interesting structurs can be observed from the basis alone. For instance basis vectors with only two reactions indicate a constant relation of $k_i/k_j = const".

Perturbance experiment
----------------------

Now, let's set up a small perturbation experiment in order to execute the differential k-cone analysis. We will use a faster version of the eigenpathways
as reference parameters and perturb the steady state value of pyruvate.

```{r}
k = 1e3*ep

# Now we generate several new steady states coming from a disturbed ethanol 
# production 

scales = 10^c(-2,-4)
d_concs = concs
d_eps = ep 
d_basis = list(kc)
for( i in 1:length(scales) ) {
	d_c = concs
	d_c["pyr"] = d_c["pyr"]*scales[i]
	d_concs = rbind(d_concs, d_c)
	basis = get_polytope_basis(S, get_ma_terms(S, d_c))
	d_basis[[i+1]] = basis
	d_eps = rbind(d_eps, eigenpathways(basis)[,1])
}
``` 

First we will have a look at the projection of the respective k-cones in a 2D space
with colors from green to red indicating a diminishing ethanol production.

```{r fig.width=6, fig.height=6}
plot_red(d_basis)
```

So we can already observe that only a small portion of the k-cone is influenced.
In order to see which parts of the basis are changed we will compare the reference
k-cone to the one obtained from the largest perturbation. We can use the dycone function `d()` to get the differnce between two basis, even if they have different numbers of basis vectors.

```{r fig.width=24, fig.height=8}
plot( d(d_basis[[3]], kc) )
```
We see that the most prominent change is an increased reaction rate for two reactions, 
which result to be the two reactions consuming Pyruvate. However, there are also some other
changes which represent alternative hypothesis what might have caused the change in
steady states. In order to investigate them in more detail there are two strategies.

First, we can evaluate the k-cones by its eigenpathways alone. This gives us some
comparative analysis over the most representative k-vector in the space.

```{r fig.width=25, fig.height=8}
barplot( d_eps, col=TRANSCOL(3), names=1:ncol(d_eps), xlab="reactions", beside=T )
```

Here we also see some other reactions popping up.
Alternatively, we can also generate hypothesis from the entire space by seeing which
reactions are changed at least by 1% across all k-cone basis vectors. This is done
with the function `hyp()`.

```{r}
h = hyp(kc, d_basis[[3]], reacts, tol=1e-2)
print(h)
```

As we can see the change in Pyruvate is explained by either up-regulating the pyruvate
consumption or (less likely) a lower production of upstream substrates which might have to act together. 
As we can see the reaction producing pyruvate is not in the list. This makes sense since it is a 
unidirectional reaction whose down-regulation would affect the steady state quite severely
in the concentration of adp and pep, but not pyruvate, thus making it an unlikely target 
when observing a lower pyruvate concentration alone.
